---
title: "GLM AND GAM"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r}

# Load required libraries
library(ggplot2)

# Load the data
data <- read.csv("telecom_customers_churn_cleaned.csv")
data$Churn <- ifelse(data$Churn == "Yes", 1, 0)
#str(data)
#colSums(is.na(data))
```


# GLM

```{r}
# Ensure variables are factors
data$Contract <- as.factor(data$Contract)
data$PaymentMethod <- as.factor(data$PaymentMethod)
data$Churn <- as.factor(data$Churn)
data$TotalCharges <- log(data$TotalCharges)
data$MonthlyCharges <- log(data$MonthlyCharges)
data$InternetService <- as.factor(data$InternetService)


# Set 'No service' as the reference level for InternetService
data$InternetService <- relevel(data$InternetService, ref = "No")
# Confirm reference level
cat("Reference level for 'Contract':", levels(data$Contract)[1], "\n")
cat("Reference level for 'InternetService':", levels(data$InternetService)[1], "\n")
# Fit the GLM with interaction terms
glm_model_interaction <- glm(Churn ~ SeniorCitizen + Contract + PaymentMethod + 
                             tenure + TotalCharges + 
                             InternetService + OnlineSecurity + TechSupport + 
                             PaperlessBilling + MultipleLines, 
                             family = binomial, data = data)
# Check the summary
summary(glm_model_interaction)

```
## So let's interpret the results;

### Fiber optic users are more likely to churn, while DSL users are less likely to churn compared to no internet service customers.

### Electronic check users are at higher risk of churn.

### Customers with higher tenure and total charges are less likely to churn.

### Longer contracts significantly reduce churn risk.

### Senior citizens are more likely to churn

### Electronic check users are more likely to churn.

### Online security increases churn

### Tech support significantly increases churn risk.

### Longer contracts significantly reduce churn.


# Did the model overfit?

```{r}
library(pROC)
# Split data into training and test sets
set.seed(123)
train_indices <- sample(1:nrow(data), 0.7 * nrow(data))
train_data <- data[train_indices, ]
test_data <- data[-train_indices, ]

# Predict on training and test data
train_preds <- predict(glm_model_interaction, train_data, type = "response")
test_preds <- predict(glm_model_interaction, test_data, type = "response")

# Calculate performance metrics
library(pROC)
train_auc <- roc(train_data$Churn, train_preds)$auc
test_auc <- roc(test_data$Churn, test_preds)$auc

cat("Train AUC:", train_auc, "\n")
cat("Test AUC:", test_auc, "\n")
```
### The Train AUC (0.847) and Test AUC (0.854) are very close, indicating that the model performs consistently on both the training and test datasets. This suggests good generalization without overfitting to the training data. 

```{r}
library(caret)

# Set up cross-validation
control <- trainControl(method = "cv", number = 10) # 10-fold cross-validation

# Train the model
cv_model <- train(Churn ~ SeniorCitizen + Contract + PaymentMethod + 
                  tenure + MonthlyCharges + TotalCharges + 
                  InternetService + InternetService:MonthlyCharges, 
                  data = data, method = "glm", family = "binomial", trControl = control)

# Check cross-validation results
print(cv_model)
```
### If there were significant overfitting, we would expect a much larger difference between training and cross-validation accuracy. So it seems there is no overfit

# GAM WITH NON-LINEAR PATTERNS

```{r}
library(mgcv)
data$TotalCharges_centered <- data$TotalCharges- mean(data$TotalCharges)
cat("Reference level for 'Contract':", levels(data$Contract)[1], "\n")
cat("Reference level for 'InternetService':", levels(data$InternetService)[1], "\n")
gam_smoothing_spline <- gam(Churn ~ SeniorCitizen + Contract + PaymentMethod + 
                             s(tenure) + s(TotalCharges_centered) + tenure +  TotalCharges +
                             InternetService + OnlineSecurity + TechSupport + 
                             PaperlessBilling + MultipleLines, 
                             family = binomial, data = data)

summary(gam_smoothing_spline)
```


## Smooth terms for tenure and TotalCharges_centered are highly significant (p < 0.001), indicating non-linear effects.

# Let's compare these models

# ROC COMPARISON

```{r}

library(pROC)

# Predictions for glm_model
pred1 <- predict(glm_model_interaction, type = "response")
roc1 <- roc(data$Churn, pred1)

pred2 <- predict(gam_smoothing_spline, type = "response")
roc2 <- roc(data$Churn, pred2)

# Plot ROC curves
plot(roc1, col = "blue", lwd = 2, main = "ROC Curve Comparison")
lines(roc2, col = "red", lwd = 2)
legend("bottomright", legend = c("glm_model_interaction", "gam_smoothing_spline"),
       col = c("blue", "red"), lwd = 2)

auc1 <- auc(roc1)
auc2 <- auc(roc2)
cat("AUC of Model glm_model_interaction:", auc1, "\n")
cat("AUC of Model gam_smoothing_spline:", auc2, "\n")
```


## Both models perform similarly in terms of AUC, but GAM slightly edges out GLM in predictive performance.


```{r}
anova(glm_model_interaction, gam_smoothing_spline, test = "Chisq")
```
## GAM explains more variance than GLM.

```{r}
AIC(glm_model_interaction, gam_smoothing_spline)
```
## GAM has a slightly lower AIC (5810.983) than GLM (5828.536), further supporting that GAM provides a better fit to the data.


# Let's plot some results


## Plot: Tenure vs. TotalCharges by Churn

```{r}
# Ensure necessary libraries are installed
if(!require(ggplot2)) install.packages("ggplot2")

# Load the library
library(ggplot2)


ggplot(data, aes(x = tenure, y = TotalCharges, color = Churn)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Relationship Between Tenure and Total Charges by Churn",
    x = "Tenure (Months)",
    y = "Total Charges"
  ) +
  scale_color_manual(values = c("blue", "red"), labels = c("No Churn", "Churn")) +
  theme_minimal()
```


## Plot: Tenure by Contract Type

```{r}

ggplot(data, aes(x = Contract, y = tenure, fill = Contract)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Tenure by Contract Type",
    x = "Contract Type",
    y = "Tenure (Months)"
  ) +
  scale_fill_manual(values = c("blue", "orange", "purple")) +
  theme_minimal()
```


## Plot: Churn by Payment Method


```{r}

ggplot(data, aes(x = PaymentMethod, fill = Churn)) +
  geom_bar(position = "fill", alpha = 0.8) +
  labs(
    title = "Proportion of Churn by Payment Method",
    x = "Payment Method",
    y = "Proportion"
  ) +
  scale_fill_manual(values = c("blue", "red"), labels = c("No Churn", "Churn")) +
  theme_minimal() +
  coord_flip()
```


## Plot: Churn Proportion by Internet Service

```{r}
# Load necessary libraries
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(dplyr)) install.packages("dplyr")

library(ggplot2)
library(dplyr)

# Create a dataset for proportions and counts
proportion_df <- data %>%
  group_by(InternetService, Churn) %>%
  summarise(Count = n()) %>%
  mutate(Total = sum(Count),
         Proportion = Count / Total)

# Plot the bar chart with counts displayed
ggplot(proportion_df, aes(x = InternetService, y = Proportion, fill = as.factor(Churn))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9), vjust = -0.2, size = 3) +
  labs(
    title = "Proportion of Churn by Internet Service with Counts",
    x = "Internet Service Type",
    y = "Proportion",
    fill = "Churn"
  ) +
  scale_fill_manual(values = c("blue", "red"), labels = c("No", "Yes")) +
  theme_minimal()
```



# Conclusion

Both models performed well in predicting churn, with GAM offering additional flexibility to capture non-linear relationships. While the non-linearity in some variables, like tenure, was not significant, GAM successfully identified a meaningful non-linear relationship with TotalCharges. The results highlight that customers are more likely to churn if they use fiber optic internet, pay via electronic checks, lack online security or tech support, or are senior citizens. Conversely, churn risk is lower for DSL users, those with higher tenure and total charges, and customers on longer-term contracts


